---
title: "Installation of R, RStudio & git"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


::: {.box-info}
`r fontawesome::fa_i("lightbulb")`&nbsp;
**Having an up-to-date environment is the first good practice. Please install the 
latest versions of `r fontawesome::fa_i("r-project")`, `RStudio`, `Pandoc`, 
`Quarto`, and `git`.**
:::



<br/>



## Installation



### `r fontawesome::fa_i("windows")` &nbsp;Windows



1. Install `r fontawesome::fa_i("r-project")`<br/>
`r fontawesome::fa_i("link")`&nbsp; [https://cran.r-project.org/](https://cran.r-project.org/)

Select _**Download R for Windows**_ `|` _**Base**_ `|` _**Download R 4.2.3 for Windows**_ 
and follow instructions.



2. Install **`RStudio Desktop`**<br/>
`r fontawesome::fa_i("link")`&nbsp; [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)

Download the version for **Windows 10/11** and follow instructions.



3. Install **`Rtools4.2`**<br/>
`r fontawesome::fa_i("link")`&nbsp; [https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html](https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html)

**`Rtools4.2`** is strongly recommended to install some `r fontawesome::fa_i("r-project")` packages 
from sources. It contains different utilities, in particular the utility `make`.

Download the 64-bit version for Windows and follow instructions.



4. Install **`git`**<br/>
`r fontawesome::fa_i("link")`&nbsp;
[https://git-scm.com/downloads](https://git-scm.com/downloads)

Download the 64-bit version for Windows and follow instructions (select the default options).



5. Install **`pandoc`**<br/>
`r fontawesome::fa_i("link")`&nbsp;
[https://pandoc.org/](https://pandoc.org/)

The `r fontawesome::fa_i("r-project")` package [**`rmarkdown`**](https://rmarkdown.rstudio.com/) 
requires the software **`pandoc`** to convert document from one format (e.g. `.Rmd`) to 
another (e.g. `.html`).

Download the latest version of **`pandoc`** (file with the extension **.msi**) at:
[https://github.com/jgm/pandoc/releases/](https://github.com/jgm/pandoc/releases/),
and follow instructions.



6. Install **`Quarto`**<br/>
`r fontawesome::fa_i("link")`&nbsp;
[https://quarto.org/](https://quarto.org/)

**`Quarto`** is a multi-language, next-generation version of R Markdown developed 
by [Posit](https://posit.co/), and includes dozens of new features and capabilities.

Download the latest version of **`Quarto`** (file with the extension **.msi**) at:
[https://quarto.org/docs/get-started/](https://quarto.org/docs/get-started/),
and follow instructions.


`r fontawesome::fa_i("hand-point-right")`&nbsp;
In addition to this software you need to install the `r fontawesome::fa_i("r-project")` package `quarto`:

```{r echo=TRUE, eval=FALSE}
# Install quarto package ----
install.packages("quarto")
```


7. Install a **`LaTeX`** distribution

If you want to convert documents from `.Rmd` to `.pdf`, you will need **`LaTeX`**. 
You can install a lightened distribution using the `r fontawesome::fa_i("r-project")`
package [**`tinytex`**](https://yihui.org/tinytex/).
Open **`RStudio`** and run: 

```{r echo=TRUE, eval=FALSE}
# Install tinytex package ----
install.packages("tinytex")

# Install LaTeX distribution ----
tinytex::install_tinytex()
```

**Important:** If you already have a **`LaTeX`** distribution, do not install `tinytex`.


8. Check installation


Open **`RStudio`** and run:

```{r echo=TRUE, eval=FALSE}
# Get R version ----
R.version.string
## "R version 4.2.3 (2023-03-15)"

# Check if git is installed ----
Sys.which("git")
##                  git
## "C:/path/to/git.exe"

# Check if Rtools is installed ----
Sys.which("make")
##                  make 
## "C:/path/to/make.exe"

# Check if Pandoc is installed ----
Sys.which("pandoc")
##              pandoc 
## "C:/path/to/Pandoc"

# Check if Quarto is installed ----
Sys.which("Quarto")
##              Quarto 
## "C:/path/to/Quarto"

# Check if LaTeX is installed ----
Sys.which("pdflatex")
##              pdflatex 
## "C:/path/to/pdflatex" 

# Install package from sources ----
install.packages("jsonlite", type = "source")
```


<br/>


::: {.box-info}
`r fontawesome::fa_i("bulleye")`&nbsp;
Go to the section [**Configuration**](#configuration) to learn how to configure `git` & `GitHub`
:::


<br/>



### `r fontawesome::fa_i("apple")` &nbsp;macOS


The easiest way to set up your environment is by using the package manager for macOS
[**`Homebrew`**](https://brew.sh). With this tool you can install (and update) your
software directly from the terminal. First you need to install the Apple software
_Command Line Tools for Xcode_.

Open a [terminal](https://support.apple.com/en-gb/guide/terminal/apd5265185d-f365-44cb-8b09-71a064a42125/mac), 
run this line and accept the license:

```sh
# Install Xcode Command Line Tools ----
sudo xcode-select --install
```

Then you can install **`Homebrew`** itself:

```sh
# Install Homebrew ----
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"

# Check installation ----
brew --version
## Homebrew 4.0.9

# Update repositories (list of available software) ----
brew update
```


Follow these lines to install your environment:

```sh
# Install utilities for R ----
brew install --cask xquartz     ## Graphical window system
brew install gfortran           ## FORTRAN compiler for GCC

# Install R ----
brew install --cask r

# Install RStudio Desktop ----
brew install --cask rstudio

# Install git ----
brew install git

# Install pandoc ----
brew install pandoc             ## Convert documents (Rmd, html, etc.)

## Install Quarto ----
brew install --cask quarto      ## A next-generation of R Markdown
```

In addition you need to install the `r fontawesome::fa_i("r-project")` package `quarto`:

```{r echo=TRUE, eval=FALSE}
install.packages("quarto")
```


If you want to convert documents from `.Rmd` to `.pdf`, you will need **`LaTeX`**. 
You can install a lightened distribution using the `r fontawesome::fa_i("r-project")`
package [**`tinytex`**](https://yihui.org/tinytex/).
Open **`RStudio`** and run: 

```{r echo=TRUE, eval=FALSE}
# Install tinytex ----
install.packages("tinytex")

# Install LaTeX ----
tinytex::install_tinytex()
```

**Important:** If you already have a **`LaTeX`** distribution, do not install `tinytex`.

Open **`RStudio`** and check your installation:

```{r echo=TRUE, eval=FALSE}
# Get R version ----
R.version.string
## "R version 4.2.3 (2023-03-15)"

# Check if git is installed ----
Sys.which("git")
##            git
## "/usr/bin/git"

# Check if Rtools is installed ----
Sys.which("make")
##            make 
## "/usr/bin/make"

# Check if Pandoc is installed ----
Sys.which("pandoc")
##                  pandoc 
## "/usr/local/bin/pandoc"

# Check if Quarto is installed ----
Sys.which("Quarto")
##                  Quarto 
## "/usr/local/bin/Quarto"

# Check if LaTeX is installed ----
Sys.which("pdflatex")
##                  pdflatex 
## "/usr/local/bin/pdflatex" 

# Install package from sources ----
install.packages("jsonlite", type = "source")
```

<br/>

`r fontawesome::fa_i("lightbulb")`&nbsp;
**To keep your system up-to-date**, run time to time:


```sh
# Update environment ----
brew update && brew upgrade --greedy
```



<br/>


::: {.box-info}
`r fontawesome::fa_i("bulleye")`&nbsp;
Go to the section [**Configuration**](#configuration) to learn how to configure `git` & `GitHub`
:::


<br/>



### `r fontawesome::fa_i("ubuntu")`&nbsp;Ubuntu


First check your version of Ubuntu in a terminal:


```sh
## Check Ubuntu version ----
cat /etc/issue
## Ubuntu 22.04 LTS
```

<br/>

| Release   | Release name    | Repository URL                                               |
|:---------:|:----------------|:-------------------------------------------------------------|
| 22.04 LTS | Jammy Jellyfish | https://cloud.r-project.org/bin/linux/ubuntu jammy-cran40/   |
| 20.04 LTS | Focal Fossa     | https://cloud.r-project.org/bin/linux/ubuntu focal-cran40/   |
| 18.04 LTS | Bionic Beaver   | https://cloud.r-project.org/bin/linux/ubuntu bionic-cran40/  |
: Table 1. List of CRAN repositories for different versions of Ubuntu

<br/>

Install `r fontawesome::fa_i("r-project")` by following these steps 
(change the URL below by the one specific to your version of Ubuntu, cf. Table 1) :

```sh
## Install APT utilities ----
sudo apt install build-essential software-properties-common wget

## Add CRAN GPG key ----
wget -qO- https://cloud.r-project.org/bin/linux/ubuntu/marutter_pubkey.asc | sudo tee -a /etc/apt/trusted.gpg.d/cran_ubuntu_key.asc

## Add CRAN repository to APT repositories list ----
sudo add-apt-repository 'deb https://cloud.r-project.org/bin/linux/ubuntu jammy-cran40/'

## Update packages list ----
sudo apt update

## Install R ----
sudo apt install r-base r-base-dev
```


<br/>


To install **`RStudio Desktop`** visit this page: 
https://posit.co/download/rstudio-desktop/, download the
`.deb` file for Ubuntu 22 (or Ubuntu 18+/Debian 10+ depending on your version) 
and install it. 

Alternatively, you can install **`RStudio Desktop`** from the terminal: 

```sh
# Download RStudio Desktop installer (Ubuntu = 22.04) ----
wget https://download1.rstudio.org/electron/jammy/amd64/rstudio-2023.03.0-386-amd64.deb

# Install RStudio Desktop ----
sudo dpkg -i rstudio-2023.03.0-386-amd64.deb

# Remove installer ----
rm rstudio-2023.03.0-386-amd64.deb
```




Now let's install [**`git`**](https://git-scm.com/downloads): 

```sh
sudo apt install git
```

And [**`Pandoc`**](https://pandoc.org/) (to convert documents with `rmarkdown`):

```sh
sudo apt install pandoc
```


Let's install the [**Quarto**](https://quarto.org/) software, a multi-language, 
next-generation version of R Markdown developed by Posit (formerly RStudio), that 
includes dozens of new features and capabilities.


```sh
# Download Quarto installer ----
wget https://github.com/quarto-dev/quarto-cli/releases/download/v1.3.302/quarto-1.3.302-linux-amd64.deb

# Install Quarto ----
sudo dpkg -i quarto-1.3.302-linux-amd64.deb

# Remove installer ----
rm quarto-1.3.302-linux-amd64.deb
```




In addition you need to install the `r fontawesome::fa_i("r-project")` package `quarto`:

```{r echo=TRUE, eval=FALSE}
install.packages("quarto")
```




If you want to convert documents from `.Rmd` to `.pdf`, you will need **`LaTeX`**. 
You can install a lightened distribution using the `r fontawesome::fa_i("r-project")`
package [**`tinytex`**](https://yihui.org/tinytex/).
Open **`RStudio`** and run: 

```{r echo=TRUE, eval=FALSE}
# Install tinytex ----
install.packages("tinytex")

# Install LaTeX ----
tinytex::install_tinytex()
```


<br />


Open **`RStudio`** and check your installation:

```{r echo=TRUE, eval=FALSE}
# Get R version ----
R.version.string
## "R version 4.2.3 (2023-03-15)"

# Check if git is installed ----
Sys.which("git")
##            git
## "/usr/bin/git"

# Check if Rtools is installed ----
Sys.which("make")
##            make 
## "/usr/bin/make"

# Check if Pandoc is installed ----
Sys.which("pandoc")
##            pandoc 
## "/usr/bin/pandoc"

# Check if Quarto is installed ----
Sys.which("Quarto")
##            Quarto 
## "/usr/bin/Quarto"

# Check if LaTeX is installed ----
Sys.which("pdflatex")
##            pdflatex 
## "/usr/bin/pdflatex" 

# Install package from sources ----
install.packages("jsonlite", type = "source")
```

<br/>


`r fontawesome::fa_i("lightbulb")`&nbsp;
**To keep your system up-to-date**, run time to time:


```sh
## Update system ----
sudo apt update && sudo apt upgrade -y
```



<br/>


::: {.box-info}
`r fontawesome::fa_i("bulleye")`&nbsp;
Go to the section [**Configuration**](#configuration) to learn how to configure `git` & `GitHub`
:::


<br/>


## Configuration {.configuration}


### `r fontawesome::fa_i("github")`&nbsp;`git`


To use **`git`**, you need to store your credentials (user name and email)
that will be added to all your commits.
Open a terminal (you can use the terminal of **RStudio**) and run:

```sh
git config --global user.name  "Jane Doe"
git config --global user.email "jane.doe@mail.com"
```

When you create a new git project the default branch will be named _master_. 
It's recommended to replace _master_ by _main_ (for more information read this
[article](https://www.zdnet.com/article/github-to-replace-master-with-alternative-term-to-avoid-slavery-references/)).

On a terminal, run:

```sh
git config --global init.defaultBranch main
```

From now when you will initialize `git` in a new project the default branch will be **main**.


<br/>


### `r fontawesome::fa_i("github")`&nbsp;GitHub SSH keys

To communicate with [GitHub](https://github.com), you can use the protocol HTTPS (default)
or the protocol SSH (recommended). The protocol SSH is a more secure method of 
logging into a remote server (GitHub). To use this protocol, you need to generate 
an SSH keys pair that consists in creating two long strings of characters: a **public key**
and a **private key**. You can place the public key on any server (for instance GitHub), 
and then connect to the server using a device (your laptop) that has access to the private key.


To generate a new SSH keys pair proceed as follow: 

0. Create an account on **GitHub** at [https://github.com/signup](https://github.com/signup)
1. Open **`RStudio`** and click on: `Tools > Global options > Git/SVN`
2. Click on **Create RSA Key** and then click on **Create** 
(you can omit the passphrase). This new SSH keys pair has been stored in `~/.ssh/`.
3. Click on **View public key** and copy it
4. Go to the page [https://github.com/settings/keys](https://github.com/settings/keys)
and click on **New SSH key**
5. Choose a name for your key (e.g. your laptop name) and paste your public key.
6. Finally click on **Add SSH key**

<br />

::: {.box-info}
`r fontawesome::fa_i("ubuntu")` &nbsp; If you are on Ubuntu you need to install 
`ssh-askpass`: `sudo apt install ssh-askpass`
:::


<br />

Let's test the SSH connection with GitHub. Open a terminal and run:

```sh
ssh -T git@github.com
```

When you first connect to GitHub (and to any remote server), you will be asked if you trust this server:

```
The authenticity of host '[92.168.45.3]:22' can't be established.
RSA key fingerprint is ...
Are you sure you want to continue connecting (yes/no/[fingerprint])?
```

Just write `yes` and press `Enter`. This will add a line in the file 
`~/.ssh/known_hosts` that lists all your trusted servers.

Congrats! You can now communicate with GitHub using the SSH protocol.

<br />

`r fontawesome::fa_i("exclamation-triangle")` &nbsp; From now, you need to
use the SSH URL when connecting to a GitHub repo.


```sh
# HTTPS URL ----
https://github.com/github_account/github_repo.git

# SSH URL (use this one) ----
git@github.com:github_account/github_repo.git
```

<br />

`r fontawesome::fa_i("exclamation-triangle")` &nbsp; 
If you lose your private SSH key you won't be able to establish
a connection with GitHub. You'll need to delete your SSH key
on GitHub and to create a new one.


<br/>


### `r fontawesome::fa_i("github")`&nbsp;GitHub PAT

If you want to use the GitHub API, essentially with the `r fontawesome::fa_i("r-project")` package
[`usethis`](https://usethis.r-lib.org/) (and [`rcompendium`](https://cran.r-project.org/package=rcompendium)),
you need to create a [GitHub Personal Access Token](https://docs.github.com/fr/authentication/keeping-your-account-and-data-secure/creating-a-personal-access-token) (PAT).

Proceed as follow:

1. Go to [https://github.com/settings/tokens](https://github.com/settings/tokens)
2. Click on **Generate new token (classic)**
3. Choose a name for your new token (e.g. your laptop name)
4. Select no expiration date
5. In the section _Select scopes_, select at least **repo** and **workflow**
6. Copy this token.

Then open **`RStudio`** and run: 

```{r echo = TRUE, eval = FALSE}
# Install usethis package ----
install.packages("usethis")

# Open ~/.Renviron file ----
usethis::edit_r_environ()
## • Modify '~/.Renviron'
## • Restart R for changes to take effect
```

In the `~/.Renviron` file, add this new line: `GITHUB_PAT='XXX'` where `XXX` is your token.
Do not forget to restart `r fontawesome::fa_i("r-project")`.

<br />

Finally let's tell `usethis` to use the SSH protocol to communicate with GitHub. 
Open **RStudio** and run:

```{r echo = TRUE, eval = FALSE}
# Open ~/.Rprofile file ----
usethis::edit_r_profile()
## • Modify '~/.Rprofile'
## • Restart R for changes to take effect
```

In the `~/.Rprofile` file, add this new line: `options(usethis.protocol = "ssh")`.
Do not forget to restart `r fontawesome::fa_i("r-project")`.



## Corrections {.appendix}


If you see mistakes or want to suggest changes, please 
[Create an issue](https://github.com/theodatasci/theodatasci.github.io/issues)
on the source repository.



## Reuse {.appendix}


The material of this website is licensed under Creative Commons Attribution 
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).
Source code is available at https://github.com/theodatasci/theodatasci.github.io/.



## Citation {.appendix}



Calcagno V, Fronhofer EA, Gounand I, Jacquet C, Kéfi C & Massol F (2023) 
Workshop FRB-CESAB & GdR TheoMoDive: Theory-driven Analysis for Ecological Data (Edition 2023). 
